Fail Faster

### 2025-03-18 paper/paper.tex
comments about bst

\begin{figure}
  \begin{tabular}{lll}
  Library    & Generator          & Average Time per Generation (ns) \\
  \bq{}         & \texttt{int\_pair}          & 70                       \\
  \bq{}         & \texttt{int\_pair\_inlined} & 35                       \\
  ScalaCheck & \texttt{intPair}            & 458                      \\
  ScalaCheck & \texttt{intPairInlined}     & 266                     
  \end{tabular}
\caption{Microbenchmarks of Generators in \bq{} (using \texttt{core\_bench} \cn) and ScalaCheck (using \texttt{jmh} \cn)}
\label{fig:overhead-explanation-perf}
\end{figure}

@@

\subsubsection{Overhead of Choice Combinators}
After the \texttt{return}, \texttt{bind}, and base-type generation combinators
(such as \texttt{gen\_int}), the most frequently used combinators are ``choice''
combinators like \texttt{weighted\_union} which randomly select between
sub-generators.

Like \texttt{return} and \texttt{bind}, these combinators incur a performance penalty
at run time.

\jwc{
Aside from the usual fact that the compiler can't ``see through'' the
abstraction boundary to specialize, combinators like union and weighted union
necessarily allocate a lot in the generation hot path.  Even without the dumb BQ
array thing, to call union and weighted union, the types mean that you have to
allocate a list.  These allocations are extremely costly, and happen every
sample from the generator. If the generator is recursive (like the tree
generator), they occur at each recursive call.
}

In practice, \texttt{weighted\_union}s is almost always\footnote{
There are some generators where \texttt{weighted\_union} is passed a list which
was itself the result of a generator: the well-typed STLC term generator used in Section~\ref{section:eval} is an example of this.
} passed an explicitly constructed list, like
\texttt{weighted\_union [(3,g); (1,g'); (2,g'')]}. 
This is because the most common use case for \texttt{weighted\_union} is
to choose between one of the different constructors of an algebraic datatype,
the options for which are always known.
The list passed to \texttt{weighted\_union} allocated at the call site, and then \emph{immediately}
destructed by the function. In principle, since the elements of the list and its length $n$ are known statically,
a compiler could unroll the loops in \texttt{weighted\_union} to depth $n$ and specialize the function to avoid allocating the list.
In practice however almost~\jwc{Once again, GHC is a notable exception here \cn{}} no compilers perform this kind of optimization.

>>>>

\begin{figure}
  \begin{tabular}{lll}
  Library    & Generator          & Average Time per Generation (ns) \\
  \bq{}         & \texttt{int\_pair}          & 70                       \\
  \bq{}         & \texttt{int\_pair\_inlined} & 35                       \\
  ScalaCheck & \texttt{intPair}            & 458                      \\
  ScalaCheck & \texttt{intPairInlined}     & 266                     
  \end{tabular}
\caption{Microbenchmarks of Generators in \bq{} (using \texttt{core\_bench} \cn) and ScalaCheck (using \texttt{jmh} \cn). Average over 10000 generations with random seeds, and a fixed size}
\label{fig:overhead-explanation-perf}
\end{figure}

@@

\subsubsection{Overhead of Choice Combinators}
After the \texttt{return}, \texttt{bind}, and base-type generation combinators
(such as \texttt{gen\_int}), the most frequently used combinators are ``choice''
combinators like \texttt{weighted\_union} which randomly select between
sub-generators.
Like \texttt{return} and \texttt{bind}, these combinators incur a performance penalty
at run time. Aside from the previously-discussed issues that compilers
cannot see through the abstraction boundary to optimize these programs,
choice combinators come with their own particular abstraction overhead.

In practice, \texttt{weighted\_union}s is almost always\footnote{
There are some generators where \texttt{weighted\_union} is passed a list which
was itself the result of a generator: the well-typed STLC term generator used in Section~\ref{section:eval} is an example of this.
} called with an explicitly constructed list, like
\texttt{weighted\_union [(w1,g1); (w2,g2); (w3,g3)]}. 
This is because the most common use case for \texttt{weighted\_union} is
to choose between one of the different constructors of an algebraic datatype,
the options for which are always known.
This list is allocated at the call site, and then \emph{immediately}
destructed by the function. In principle, since the elements of the list and its length $n$ are known statically,
a compiler could unroll the loops in \texttt{weighted\_union} to depth $n$ and specialize the function to avoid allocating the list.
Unfortunately, almost~\jwc{Once again, GHC is a notable exception here \cn{}} no compilers perform this kind of optimization.
This allocation (or rather, its tendency to cause GC pauses) --- as well as the cost of running the code 
to traverse arbitrary lists compared to unrolled loops --- has a significant impact on 
performance.

Last, many PBT libraries --- including both \bq and ScalaCheck --- implement their \texttt{weighted\_union} combinators
in ways that are asymptotically more efficient but slower in practice than the naive algorithm.
Weighted union works by the Fitness Proportionate Selection algorithm \cn{}, which (1) samples a number $r$
between $0$ and the sum of the weights, and then (2) finds the first generator for which the cumulative
weights in the list before it exceeds $r$. This second part can be accomplished in $O(\log_2 n)$ time by
binary searching the list. However, since the lists are short in practice, a linear scan is almost always faster \cn{}.
Moreover, both \bq{} and ScalaCheck allocate auxiliary data structures (an array in \bq{} and a BST in ScalaCheck) to
perform this search, which incurs further run time overhead.
